---
title: "Script to merge data"
author: "Jaakko Koskenniemi"
output: 
  html_document:
    number_sections: true
date: '2022-24-04'
editor_options: 
  chunk_output_type: console
---

```{r}
pacman::p_load(tidyverse, rio, here, TwoSampleMR)
here::i_am("analyses/Harmonize.Rmd")
```


This script merges the annotation and protein data. 

GWAS sumstat of type 1 diabetes was obtained from the study by Chiou et al. 
(Nature 2021, https://doi.org/10.1038/s41586-021-03552-w). GWAS sumstat for 
protein levels of IL2RG was obtained from the study by Ferkingstad et al. 
(Nature 2021, https://doi.org/10.1038/s41588-021-00978-w).

Variants within 1 megabase before the start site of IL2RG and 1 megabase 
after the stop site are included. See '/code/gather.R' on how the the 
variants within this area were filtered in each GWAS. The gather-script 
was run in the high performance computing cluster using the 
'run_gather.sh' script. The output of '/code/Gather.R' is saved in 
'/data/cis_sumstats/', and for each possible target gene
there are three files: 

1. *_T1D_data.rds: GWAS sumstat for the risk of T1D (Chiou et al. 2021)
2. *_prot_data.rds: GWAS sumstat of serum protein level (Ferkingstad et al. 2021)
3. *_anno_data.rds: further annotation for GWAS of serum protein level  
Where * is a gene of interest (IL2RA, IL2RG, IL2RG, IL6ST, IL6R, etc.)

except for IL6R for which 
IL6R_prot_data.rds is substituted with IL6R_crp_data.rds

```{r}
#add here a figure that shows the project workflow
```


# pQTL data

## Import pQTL data

```{r}
IL2RG_anno  <- import("data/cis_sumstats/IL2RG_anno.rds")
IL2RG_prot  <- import("data/cis_sumstats/IL2RG_prot.rds")
IL12B_anno  <- import("data/cis_sumstats/IL12B_anno.rds")  
IL12B_prot  <- import("data/cis_sumstats/IL12B_prot.rds")  
IL2RA_anno  <- import("data/cis_sumstats/IL2RA_anno.rds")  
IL2RA_prot  <- import("data/cis_sumstats/IL2RA_prot.rds")  
IL2RB_prot  <- import("data/cis_sumstats/IL2RB_prot.rds")  
IL2RB_anno  <- import("data/cis_sumstats/IL2RB_anno.rds")  
IL6R_anno   <- import("data/cis_sumstats/IL6R_anno.rds")  
IL6R_prot   <- import("data/cis_sumstats/IL6R_prot.rds")
IL6ST_anno  <- import("data/cis_sumstats/IL6ST_anno.rds")  
IL6ST_prot  <- import("data/cis_sumstats/IL6ST_prot.rds")
JAK2_anno   <- import("data/cis_sumstats/JAK2_anno.rds")  
JAK2_prot   <- import("data/cis_sumstats/JAK2_prot.rds")  
TYK2_anno   <- import("data/cis_sumstats/TYK2_anno.rds")
TYK2_prot   <- import("data/cis_sumstats/TYK2_prot.rds")
```


## Merge protein data with annotations

Ferkingstad et al. state in their README file  

>A subset of the variants in the summary statistics files should be excluded 
due to quality issues. These variants are listed in a separate Excluded variants 
file (assocvariants.excluded.txt.gz). The Extra annotation file 
(assocvariants.annotated.txt.gz) does not include these variants.

Therefore in the join process, only the data available in the annotation (IL2RG_anno)
is used (hence, left_join)

```{r}
IL12B_prot_anno  <- left_join(IL12B_prot, IL12B_anno, by = c("Name", "Chrom", "Pos"))
IL2RA_prot_anno  <- left_join(IL2RA_prot, IL2RA_anno, by = c("Name", "Chrom", "Pos"))
IL2RB_prot_anno  <- left_join(IL2RB_prot, IL2RB_anno, by = c("Name", "Chrom", "Pos"))
IL2RG_prot_anno  <- left_join(IL2RG_prot, IL2RG_anno, by = c("Name", "Chrom", "Pos"))
IL6R_prot_anno   <- left_join(IL6R_prot,  IL6R_anno, by = c("Name", "Chrom", "Pos"))
IL6ST_prot_anno  <- left_join(IL6ST_prot, IL6ST_anno, by = c("Name", "Chrom", "Pos"))
JAK2_prot_anno   <- left_join(JAK2_prot,  JAK2_anno, by = c("Name", "Chrom", "Pos"))
TYK2_prot_anno   <- left_join(TYK2_prot,  TYK2_anno, by = c("Name", "Chrom", "Pos"))

rm("IL12B_anno", "IL12B_prot", 
   "IL2RA_anno", "IL2RA_prot", "IL2RB_anno", "IL2RB_prot",  
   "IL2RG_anno", "IL2RG_prot", "IL6R_anno", "IL6R_prot", 
   "IL6ST_anno", "IL6ST_prot", "JAK2_anno", "JAK2_prot", 
   "TYK2_anno", "TYK2_prot")
```


## Sort conflicts

Further in the README file of Ferkingstad et al.  

> Also, the summary statistics files sometimes incorrectly have effectAllele=otherAllele for multiallelic variants.
In these cases the effectAllele is correct, but the otherAllele should be '!', meaning that the effectAllele is tested against the other (two or more) alleles (using the '!' sign as shorthand for 'not effectAllele').
This has been corrected in the file Extra annotation file (assocvariants.annotated.txt.gz).

Therefore, multiallelic variants were removed.
```{r}
IL12B_prot_anno  <- IL12B_prot_anno  %>% filter(otherAllele.y != "!")
IL2RA_prot_anno  <- IL2RA_prot_anno  %>% filter(otherAllele.y != "!")
IL2RB_prot_anno  <- IL2RB_prot_anno  %>% filter(otherAllele.y != "!")
IL2RG_prot_anno  <- IL2RG_prot_anno  %>% filter(otherAllele.y != "!")
IL6R_prot_anno   <- IL6ST_prot_anno  %>% filter(otherAllele.y != "!")
IL6ST_prot_anno  <- IL6ST_prot_anno  %>% filter(otherAllele.y != "!")
JAK2_prot_anno   <- JAK2_prot_anno   %>% filter(otherAllele.y != "!")
TYK2_prot_anno   <- TYK2_prot_anno   %>% filter(otherAllele.y != "!")

```

After removal of multiallelic variants there were no conflicts 
between effectAllele.x and effectAllele.y or between 
otherAllele.x and otherAllele.y (results suppressed,
please see "analyses/Harmonize.Rmd") 

```{r, results = 'hide'}

#Following function is simply to avoid repetition
check_conflicts_ea_oa <- function(data, effectAllele.x, effectAllele.y) {
  filter(data, effectAllele.x != effectAllele.y | 
           is.na(effectAllele.x) & !is.na(effectAllele.y) |
           !is.na(effectAllele.x) &  is.na(effectAllele.y) | 
           otherAllele.x != otherAllele.y | 
            is.na(otherAllele.x) & !is.na(otherAllele.y) |
           !is.na(otherAllele.x) &  is.na(otherAllele.y))
}

#List those with conflicts in effectAllele or otherAllele

IL12B_prot_anno  %>% check_conflicts_ea_oa() #0 conflicts
IL2RA_prot_anno  %>% check_conflicts_ea_oa() #0 conflicts
IL2RB_prot_anno  %>% check_conflicts_ea_oa() #0 conflicts
IL2RG_prot_anno  %>% check_conflicts_ea_oa() #0 conflicts
IL6R_prot_anno   %>% check_conflicts_ea_oa() #0 conflicts
IL6ST_prot_anno  %>% check_conflicts_ea_oa() #0 conflicts
JAK2_prot_anno   %>% check_conflicts_ea_oa() #0 conflicts
TYK2_prot_anno   %>% check_conflicts_ea_oa() #0 conflicts

```

After this, the remaining discrepancies between the variables
obtained from annotation and protein data files were because those
variants without an rsid were listed as "." in the annotation file
and as "NA" in the main GWAS sumstat file (please see 
'/analyses/Harmonize.Rmd')

```{r, results = 'hide'}
#List subjects which had conflicting rsid information and show
#which entries there are for these
check_conflicts_rsid <- function(data, effectAllele.x, effectAllele.y) {
  filter(data, rsids.x != rsids.y | 
           is.na(rsids.x) & !is.na(rsids.y) |
           !is.na(rsids.x) &  is.na(rsids.y)) %>%
    mutate(rsids.x = factor(rsids.x),
         rsids.y = factor(rsids.y)) %>%
  select(rsids.x, rsids.y) %>%
  summary()
}

#All the conflicts in rsid.x and rsid.y are because rsid.x is NA and rsid.y is "."
IL12B_prot_anno  %>% check_conflicts_rsid() #0 conflicts
IL2RA_prot_anno  %>% check_conflicts_rsid() #0 conflicts
IL2RB_prot_anno  %>% check_conflicts_rsid() #0 conflicts
IL2RG_prot_anno  %>% check_conflicts_rsid() #0 conflicts
IL6R_prot_anno   %>% check_conflicts_rsid() #0 conflicts
IL6ST_prot_anno  %>% check_conflicts_rsid() #0 conflicts
JAK2_prot_anno   %>% check_conflicts_rsid() #0 conflicts
TYK2_prot_anno   %>% check_conflicts_rsid() #0 conflicts

```

Thus, only those values variables with .x in their name are preserved. 
Here, we also remove the remaining multiallelic SNPs 
(where effect allele equals the other allele.)

```{r}
#Discard all variables that end with an ".y"
IL12B_prot_anno   <- IL12B_prot_anno  %>% select(!ends_with(".y"))
IL2RA_prot_anno   <- IL2RA_prot_anno  %>% select(!ends_with(".y"))
IL2RB_prot_anno   <- IL2RB_prot_anno  %>% select(!ends_with(".y"))
IL2RG_prot_anno   <- IL2RG_prot_anno  %>% select(!ends_with(".y"))
IL6R_prot_anno    <- IL6ST_prot_anno  %>% select(!ends_with(".y"))
IL6ST_prot_anno   <- IL6ST_prot_anno  %>% select(!ends_with(".y"))
JAK2_prot_anno    <- JAK2_prot_anno   %>% select(!ends_with(".y"))
TYK2_prot_anno    <- TYK2_prot_anno   %>% select(!ends_with(".y"))

#Remove ".x" from variable names
IL12B_prot_anno   <- IL12B_prot_anno  %>% rename_with(~gsub(".x", "", .x, fixed = TRUE)) 
IL2RA_prot_anno   <- IL2RA_prot_anno  %>% rename_with(~gsub(".x", "", .x, fixed = TRUE)) 
IL2RB_prot_anno   <- IL2RB_prot_anno  %>% rename_with(~gsub(".x", "", .x, fixed = TRUE)) 
IL2RG_prot_anno   <- IL2RG_prot_anno  %>% rename_with(~gsub(".x", "", .x, fixed = TRUE)) 
IL6ST_prot_anno   <- IL6ST_prot_anno  %>% rename_with(~gsub(".x", "", .x, fixed = TRUE)) 
IL6R_prot_anno    <- IL6R_prot_anno   %>% rename_with(~gsub(".x", "", .x, fixed = TRUE)) 
JAK2_prot_anno    <- JAK2_prot_anno   %>% rename_with(~gsub(".x", "", .x, fixed = TRUE)) 
TYK2_prot_anno    <- TYK2_prot_anno   %>% rename_with(~gsub(".x", "", .x, fixed = TRUE)) 
```


```{r, results='hide'}
#Sanity checks
colnames(IL2RG_prot_anno)   
colnames(IL12B_prot_anno)
colnames(IL2RA_prot_anno)
colnames(IL2RB_prot_anno)
colnames(IL2RG_prot_anno)
colnames(IL6ST_prot_anno)
colnames(IL6R_prot_anno)
colnames(JAK2_prot_anno)
colnames(TYK2_prot_anno)
```

All data files seem to have entries without rsids (suppressed,
please see '/analyses/Harmonize.Rmd'). They are now removed

```{r, results = 'hide'}
anyNA(IL2RG_prot_anno$rsids)   
anyNA(IL12B_prot_anno$rsids)
anyNA(IL2RA_prot_anno$rsids)
anyNA(IL2RB_prot_anno$rsids)
anyNA(IL2RG_prot_anno$rsids)
anyNA(IL6ST_prot_anno$rsids)
anyNA(IL6R_prot_anno$rsids)
anyNA(JAK2_prot_anno$rsids)
anyNA(TYK2_prot_anno$rsids)

IL2RG_prot_anno   <- IL2RG_prot_anno  %>% filter(!is.na(rsids))
IL12B_prot_anno   <- IL12B_prot_anno  %>% filter(!is.na(rsids))
IL2RA_prot_anno   <- IL2RA_prot_anno  %>% filter(!is.na(rsids))
IL2RB_prot_anno   <- IL2RB_prot_anno  %>% filter(!is.na(rsids))
IL2RG_prot_anno   <- IL2RG_prot_anno  %>% filter(!is.na(rsids))
IL6R_prot_anno    <- IL6ST_prot_anno  %>% filter(!is.na(rsids))
IL6ST_prot_anno   <- IL6ST_prot_anno  %>% filter(!is.na(rsids))
JAK2_prot_anno    <- JAK2_prot_anno   %>% filter(!is.na(rsids))
TYK2_prot_anno    <- TYK2_prot_anno   %>% filter(!is.na(rsids))
```


There were no duplicated rsids in data after exclusion of missing 
rsids (suppressed, please see /analyses/Harmonize.Rmd)

```{r, results='hide'}
#Sanity check
anyDuplicated(IL2RG_prot_anno$rsids)   
anyDuplicated(IL12B_prot_anno$rsids)
anyDuplicated(IL2RA_prot_anno$rsids)
anyDuplicated(IL2RB_prot_anno$rsids)
anyDuplicated(IL2RG_prot_anno$rsids)
anyDuplicated(IL6ST_prot_anno$rsids)
anyDuplicated(IL6R_prot_anno$rsids)
anyDuplicated(JAK2_prot_anno$rsids)
anyDuplicated(TYK2_prot_anno$rsids)
```


## Write data in TwoSampleMR format

Data are written in TwoSampleMR format so that they can be easily opened
in the main analyses scripts.

Rename files
```{r}
rename_for_TwoSampleMR_e  <- function(data, rsids, Beta, SE, effectAlleleFreq,
                                    effectAllele, otherAllele,
                                    Pval, Pos, N) {
  data %>%
    rename(SNP = rsids, beta = Beta, se = SE, eaf = effectAlleleFreq,
         effect_allele = effectAllele, other_allele = otherAllele,
         pval = Pval, pos = Pos) %>%
    mutate(samplesize = N) 
  }

#Rename
IL12B_prot_anno   <- IL12B_prot_anno  %>% rename_for_TwoSampleMR_e()
IL2RA_prot_anno   <- IL2RA_prot_anno  %>% rename_for_TwoSampleMR_e()
IL2RB_prot_anno   <- IL2RB_prot_anno  %>% rename_for_TwoSampleMR_e()
IL2RG_prot_anno   <- IL2RG_prot_anno  %>% rename_for_TwoSampleMR_e()
IL6R_prot_anno    <- IL6R_prot_anno   %>% rename_for_TwoSampleMR_e()
IL6ST_prot_anno   <- IL6ST_prot_anno  %>% rename_for_TwoSampleMR_e()
JAK2_prot_anno    <- JAK2_prot_anno   %>% rename_for_TwoSampleMR_e()
TYK2_prot_anno    <- TYK2_prot_anno   %>% rename_for_TwoSampleMR_e()

#Add phenotype information
IL12B_prot_anno   <- IL12B_prot_anno  %>%  
  mutate(Phenotype = "Serum IL12B level")
IL2RA_prot_anno   <- IL2RA_prot_anno  %>%  
  mutate(Phenotype = "Serum IL2RA level")
IL2RB_prot_anno   <- IL2RB_prot_anno  %>%  
  mutate(Phenotype = "Serum IL2RB level")
IL2RG_prot_anno   <- IL2RG_prot_anno  %>%  
  mutate(Phenotype = "Serum IL2RG level")
IL6R_prot_anno    <- IL6R_prot_anno  %>%
  mutate(Phenotype = "Serum IL6R level")
IL6ST_prot_anno   <- IL6ST_prot_anno  %>%
  mutate(Phenotype = "Serum IL6ST level")
JAK2_prot_anno    <- JAK2_prot_anno   %>%  
  mutate(Phenotype = "Serum JAK2 level")
TYK2_prot_anno    <- TYK2_prot_anno   %>%  
  mutate(Phenotype = "Serum TYK2 level")
```


Export filest to a CSV file with default parameters of 
TwoSampleMR::read_exposure_data()
```{r}
#Export
IL12B_prot_anno  %>% 
  export("data/export/IL12B_prot_anno_TwoSampleMR.csv")
IL2RA_prot_anno  %>% 
  export("data/export/IL2RA_prot_anno_TwoSampleMR.csv")
IL2RB_prot_anno  %>% 
  export("data/export/IL2RB_prot_anno_TwoSampleMR.csv")
IL2RG_prot_anno  %>% 
  export("data/export/IL2RG_prot_anno_TwoSampleMR.csv")
IL6R_prot_anno  %>%
  export("data/export/IL6R_prot_anno_TwoSampleMR.csv")
IL6ST_prot_anno  %>%
  export("data/export/IL6ST_prot_anno_TwoSampleMR.csv")
JAK2_prot_anno   %>% 
  export("data/export/JAK2_prot_anno_TwoSampleMR.csv")
TYK2_prot_anno   %>% 
  export("data/export/TYK2_prot_anno_TwoSampleMR.csv")
```

```{r, results = 'hide'}

#Sanity check: no warnings
# read_exposure_data("data/export/IL12B_prot_anno_TwoSampleMR.csv", sep = ",")
# read_exposure_data("data/export/IL2RA_prot_anno_TwoSampleMR.csv", sep = ",")
# read_exposure_data("data/export/IL2RB_prot_anno_TwoSampleMR.csv", sep = ",")
# read_exposure_data("data/export/IL2RG_prot_anno_TwoSampleMR.csv", sep = ",")
# read_exposure_data("data/export/IL6ST_prot_anno_TwoSampleMR.csv", sep = ",")
# read_exposure_data("data/export/IL6R_prot_anno_TwoSampleMR.csv", sep = ",")
# read_exposure_data("data/export/JAK2_prot_anno_TwoSampleMR.csv", sep = ",")
# read_exposure_data("data/export/TYK2_prot_anno_TwoSampleMR.csv", sep = ",")
```

# IL6 CRP data
## Write IL6R_crp in TwoSampleMR format

```{r}
IL6R_crp   <- import("data/cis_sumstats/IL6R_crp.rds")  

# colnames(IL6R_crp)
IL6R_crp <- IL6R_crp %>%
  select(-other_allele, -effect_allele, -beta) %>%
  rename(SNP = hm_rsid, beta = hm_beta, se = standard_error, 
           eaf = hm_effect_allele_frequency,
           effect_allele = hm_effect_allele, 
           other_allele = hm_other_allele,
           pval = p_value, pos = hm_pos) %>%
  mutate(pval = as.numeric(pval),
         eaf = as.numeric(eaf),
         samplesize = 575531,
         Phenotype = "Serum C-reactive protein levels") 

IL6R_crp  %>% 
  export("data/export/IL6R_crp_TwoSampleMR.csv")
```


#eQTL data

## Import data
```{r}
library(data.table)
library(dtplyr)

#Read eqtl sumstat
IL2RA_eqtl  <- import("data/cis_eqtl_sumstats/eqtl_IL2RA-Gather.rds") %>% as_tibble()
IL2RG_eqtl  <- import("data/cis_eqtl_sumstats/eqtl_IL2RG-Gather.rds") %>% as_tibble()
IL2RB_eqtl  <- import("data/cis_eqtl_sumstats/eqtl_IL2RB-Gather.rds") %>% as_tibble()
IL6R_eqtl   <- import("data/cis_eqtl_sumstats/eqtl_IL6R-Gather.rds") %>% as_tibble()
JAK1_eqtl   <- import("data/cis_eqtl_sumstats/eqtl_JAK1-Gather.rds") %>% as_tibble()
JAK2_eqtl   <- import("data/cis_eqtl_sumstats/eqtl_JAK2-Gather.rds") %>% as_tibble()
JAK3_eqtl   <- import("data/cis_eqtl_sumstats/eqtl_JAK3-Gather.rds") %>% as_tibble()
TYK2_eqtl   <- import("data/cis_eqtl_sumstats/eqtl_TYK2-Gather.rds") %>% as_tibble()
IFNAR2_eqtl <- import("data/cis_eqtl_sumstats/eqtl_IFNAR2-Gather.rds") %>% as_tibble()

#IFNAR2 had a lot of duplicated SNPs, and on closer inspection, there are two genes with different Ensmbl-symbols under IFNAR2 (ENSG00000159110 and ENSG00000249624). The former is IFNAR2 and the other codes a "novel gene"
IFNAR2_eqtl$Gene %>% factor() %>% levels()

#None of the other genes had two Ensemble codes
IL2RA_eqtl$Gene %>% factor() %>% levels()
IL2RG_eqtl$Gene %>% factor() %>% levels()
IL2RB_eqtl$Gene %>% factor() %>% levels()
IL6R_eqtl$Gene %>% factor() %>% levels()
JAK1_eqtl$Gene %>% factor() %>% levels()
JAK2_eqtl$Gene %>% factor() %>% levels()
JAK3_eqtl$Gene %>% factor() %>% levels()
TYK2_eqtl$Gene %>% factor() %>% levels()

#Therefore, all the entries with "Gene == ENSG00000249624" were removed
IFNAR2_eqtl <- IFNAR2_eqtl %>% filter(Gene == "ENSG00000159110")


names_eqtl <- c("IFNAR2_eqtl", 
"IL2RA_eqtl", 
"IL2RG_eqtl", 
"IL2RB_eqtl", 
"IL6R_eqtl", 
"JAK1_eqtl", 
"JAK2_eqtl", 
"JAK3_eqtl", 
"TYK2_eqtl")

```


According to the readme files of eQTLGEN phase I 
(data/REAMDE_cis_eqtl.txt and data/README_cis_eqtl_allele_frequency.txt),
```{r, echo = FALSE}
c("Pvalue", "P-value",
  "SNP", "SNP rs ID",
  "SNPChr", "SNP chromosome",
  "SNPPos", "SNP position",
  "AssessedAllele", "Assessed allele, the Z-score refers to this allele",
  "OtherAllele", "Not assessed allele",
  "Zscore", "Z-score",
  "Gene", "ENSG name (Ensembl v71) of the eQTL gene",
  "GeneSymbol", "HGNC name of the gene",
  "GeneChr", "Gene chromosome",
  "GenePos", "Centre of gene position",
  "NrCohorts", "Total number of cohorts where this SNP-gene combination was tested",
  "NrSamples", "Total number of samples where this SNP-gene combination was tested",
  "FDR", "False discovery rate estimated based on permutations",
  "BonferroniP", "P-value after Bonferroni correction",
  "SNP", "SNP rs ID",
  "hg19_chr", "chr number",
  "hg19_pos", "SNP position (hg19)",
  "AlleleA", "Other allele",
  "AlleleB", "Assessed allele",
  "allA_total", "Total allele count of genotype AA",
  "allAB_total", "Total allele count of genotype AB",
  "allB_total", "Total allele count of genotype BB",
  "AlleleB_all", "Allele frequency of assessed allele") %>%
  matrix(byrow =T, ncol = 2) %>% 
  cbind(c(rep("README_cis_eqtl.txt", 15), rep("README_cis_eqtl_AF.txt", 9))) %>%
  knitr::kable(caption = "sources: README_cis_eqtl_AF.txt = https://molgenis26.gcc.rug.nl/downloads/eqtlgen/cis-eqtl/README_allele_frequency,
  README_cis_eqtl.txt = https://molgenis26.gcc.rug.nl/downloads/eqtlgen/cis-eqtl/README_cis")
```


There is some disagreement between the readme and the original file. 
E.g. there are 1,053 subjects near IL2RB which for which 'AssessedAllele'
is not equal to 'AlleleB'. 

```{r}
colnames(IL2RB_eqtl)

get_n_conflicts  <- function(data, AlleleB=AlleleB, AssessedAllele=AssessedAllele) {
  data %>%
  filter(AlleleB != AssessedAllele) %>% 
  pull(SNP) %>% length()
}

table_conflicts_eqtl <- c("IFNAR2", get_n_conflicts(IFNAR2_eqtl),
  "IL2RA", get_n_conflicts(IL2RA_eqtl),
  "IL2RG", get_n_conflicts(IL2RG_eqtl),
  "IL2RB", get_n_conflicts(IL2RB_eqtl),
  "IL6R", get_n_conflicts(IL6R_eqtl),
  "JAK1", get_n_conflicts(JAK1_eqtl),
  "JAK2", get_n_conflicts(JAK2_eqtl),
  "JAK3", get_n_conflicts(JAK3_eqtl),
  "TYK2", get_n_conflicts(TYK2_eqtl)) %>%
  matrix(ncol = 2, byrow = TRUE)

colnames(table_conflicts_eqtl) <- c("EQTL", "n conflicts")

knitr::kable(table_conflicts_eqtl)
```



```{r}
summarize_agreement <- function(data, AlleleB=AlleleB, 
                           AssessedAllele=AssessedAllele, 
                           AlleleB_all=AlleleB_all) {
  data %>%
  mutate(AlleleB_vs_AssessedAllele = 
           ifelse(AlleleB == AssessedAllele, "AlleleB == AssessedAllele", 
                  "AlleleB != AssessedAllele")) %>%
  group_by(AlleleB_vs_AssessedAllele) %>%
  summarise(p_AF_m40  = round(sum(AlleleB_all < 0.40) / length(AlleleB_all), 4),
            p_AF_4050 = round(sum(AlleleB_all >= 0.40 & AlleleB_all < 0.50)  /
                                length(AlleleB_all), 4),
            p_AF_5060 = round(sum(AlleleB_all > 0.50 & AlleleB_all <= 0.60)  /
                                length(AlleleB_all), 4),
            p_AF_p60  = round(sum(AlleleB_all > 0.60) / length(AlleleB_all), 4),
            p_AF_50  = round(sum(AlleleB_all == 0.50) / length(AlleleB_all), 4),
            n_AF_NA = sum(is.na(AlleleB_all)))
}
```


In a remarkably stable pattern between files, AlleleB == Assessed allele when
whenAF < 0.50 and AlleleB != Assessed allele when AF > 0.50.
```{r}
summarize_agreement(IFNAR2_eqtl) %>% 
  knitr::kable(caption = "IFNAR2_eqtl")
summarize_agreement(IL2RA_eqtl) %>% 
  knitr::kable(caption = "IL2RA_eqtl")
summarize_agreement(IL2RB_eqtl) %>% 
  knitr::kable(caption = "IL2RB_eqtl")
summarize_agreement(IL6R_eqtl) %>% 
  knitr::kable(caption = "IL6R_eqtl")
summarize_agreement(JAK1_eqtl) %>% 
  knitr::kable(caption = "JAK1_eqtl")
summarize_agreement(JAK2_eqtl) %>% 
  knitr::kable(caption = "JAK2_eqtl")
summarize_agreement(JAK3_eqtl) %>% 
  knitr::kable(caption = "JAK3_eqtl")
summarize_agreement(TYK2_eqtl) %>% 
  knitr::kable(caption = "TYK2_eqtl")
```

From this, it is taken that AF is given per effect alleles mentioned 
in README_cis_eqtl_AF.txt. Therefore, EAFs are harmonized to align with
the variable 'AssessedAllele'

```{r}
IFNAR2_eqtl <- IFNAR2_eqtl %>% rename(AF_B = AlleleB_all)
IL2RA_eqtl  <- IL2RA_eqtl  %>% rename(AF_B = AlleleB_all)
IL2RG_eqtl  <- IL2RG_eqtl  %>% rename(AF_B = AlleleB_all)
IL2RB_eqtl  <- IL2RB_eqtl  %>% rename(AF_B = AlleleB_all)
IL6R_eqtl   <- IL6R_eqtl   %>% rename(AF_B = AlleleB_all)
JAK1_eqtl   <- JAK1_eqtl   %>% rename(AF_B = AlleleB_all)
JAK2_eqtl   <- JAK2_eqtl   %>% rename(AF_B = AlleleB_all)
JAK3_eqtl   <- JAK3_eqtl   %>% rename(AF_B = AlleleB_all)
TYK2_eqtl   <- TYK2_eqtl   %>% rename(AF_B = AlleleB_all)

IFNAR2_eqtl <- IFNAR2_eqtl %>%
  mutate(EAF = ifelse(AlleleB == AssessedAllele, AF_B, 1 - AF_B))
IL2RA_eqtl  <- IL2RA_eqtl %>%
  mutate(EAF = ifelse(AlleleB == AssessedAllele, AF_B, 1 - AF_B))
IL2RG_eqtl  <- IL2RG_eqtl %>%
  mutate(EAF = ifelse(AlleleB == AssessedAllele, AF_B, 1 - AF_B))
IL2RB_eqtl  <- IL2RB_eqtl %>%
  mutate(EAF = ifelse(AlleleB == AssessedAllele, AF_B, 1 - AF_B))
IL6R_eqtl   <- IL6R_eqtl %>%
  mutate(EAF = ifelse(AlleleB == AssessedAllele, AF_B, 1 - AF_B))
JAK1_eqtl   <- JAK1_eqtl %>%
  mutate(EAF = ifelse(AlleleB == AssessedAllele, AF_B, 1 - AF_B))
JAK2_eqtl   <- JAK2_eqtl %>%
  mutate(EAF = ifelse(AlleleB == AssessedAllele, AF_B, 1 - AF_B))
JAK3_eqtl   <- JAK3_eqtl %>%
  mutate(EAF = ifelse(AlleleB == AssessedAllele, AF_B, 1 - AF_B))
TYK2_eqtl   <- TYK2_eqtl %>%
  mutate(EAF = ifelse(AlleleB == AssessedAllele, AF_B, 1 - AF_B))


IFNAR2_eqtl <- IFNAR2_eqtl %>%
  mutate(MAF = ifelse(EAF < 0.5,  EAF, 1 - EAF))
IL2RA_eqtl  <- IL2RA_eqtl %>%
  mutate(MAF = ifelse(EAF < 0.5,  EAF, 1 - EAF))
IL2RG_eqtl  <- IL2RG_eqtl %>%
  mutate(MAF = ifelse(EAF < 0.5,  EAF, 1 - EAF))
IL2RB_eqtl  <- IL2RB_eqtl %>%
  mutate(MAF = ifelse(EAF < 0.5,  EAF, 1 - EAF))
IL6R_eqtl   <- IL6R_eqtl %>%
  mutate(MAF = ifelse(EAF < 0.5,  EAF, 1 - EAF))
JAK1_eqtl   <- JAK1_eqtl %>%
  mutate(MAF = ifelse(EAF < 0.5,  EAF, 1 - EAF))
JAK2_eqtl   <- JAK2_eqtl %>%
  mutate(MAF = ifelse(EAF < 0.5,  EAF, 1 - EAF))
JAK3_eqtl   <- JAK3_eqtl %>%
  mutate(MAF = ifelse(EAF < 0.5,  EAF, 1 - EAF))
TYK2_eqtl   <- TYK2_eqtl %>%
  mutate(MAF = ifelse(EAF < 0.5,  EAF, 1 - EAF))
```




```{r}
rename_eqtl_data <- function(data, Phenotype, Zscore, EAF, AssessedAllele, OtherAllele,
                              Pvalue, SNPPos, NrSamples) {
   data %>%
  rename(beta = Zscore, 
         eaf = EAF,
         effect_allele = AssessedAllele, 
         other_allele = OtherAllele,
         pval = Pvalue, 
         pos = SNPPos,
         samplesize = NrSamples) %>%
  mutate(pval = as.numeric(pval),
         eaf = as.numeric(eaf),
         Phenotype = Phenotype,
         se = NA) 
 }

IFNAR2_eqtl %>%  rename_eqtl_data(Phenotype = "Whole blood IFNAR2 mRNA") %>% export("data/export/IFNAR2_eqtl_TwoSampleMR.csv")
IL2RA_eqtl  %>%  rename_eqtl_data(Phenotype = "Whole blood IL2RA mRNA") %>% export("data/export/IL2RA_eqtl_TwoSampleMR.csv")
IL2RG_eqtl  %>%  rename_eqtl_data(Phenotype = "Whole blood IL2RG mRNA") %>% export("data/export/IL2RG_eqtl_TwoSampleMR.csv")
IL2RB_eqtl  %>%  rename_eqtl_data(Phenotype = "Whole blood IL2RB mRNA") %>% export("data/export/IL2RB_eqtl_TwoSampleMR.csv")
IL6R_eqtl   %>%  rename_eqtl_data(Phenotype = "Whole blood IL6L mRNA") %>% export("data/export/IL6R_eqtl_TwoSampleMR.csv")
JAK1_eqtl   %>%  rename_eqtl_data(Phenotype = "Whole blood JAK1 mRNA") %>% export("data/export/JAK1_eqtl_TwoSampleMR.csv")
JAK2_eqtl   %>%  rename_eqtl_data(Phenotype = "Whole blood JAK2 mRNA") %>% export("data/export/JAK2_eqtl_TwoSampleMR.csv")
JAK3_eqtl   %>%  rename_eqtl_data(Phenotype = "Whole blood JAK3 mRNA") %>% export("data/export/JAK3_eqtl_TwoSampleMR.csv")
TYK2_eqtl   %>%  rename_eqtl_data(Phenotype = "Whole blood TYK2 mRNA") %>% export("data/export/TYK2_eqtl_TwoSampleMR.csv")

```




# Outcome data

## Import outcome data


```{r}
IL12B_T1D   <- import("data/cis_sumstats/IL12B_T1D.rds") %>% as_tibble()  
TYK2_T1D    <- import("data/cis_sumstats/TYK2_T1D.rds")  %>% as_tibble()
JAK1_T1D    <- import("data/cis_sumstats/JAK1_T1D.rds")  %>% as_tibble()
JAK2_T1D    <- import("data/cis_sumstats/JAK2_T1D.rds")  %>% as_tibble()
JAK3_T1D    <- import("data/cis_sumstats/JAK3_T1D.rds")  %>% as_tibble()
IL6ST_T1D   <- import("data/cis_sumstats/IL6ST_T1D.rds") %>% as_tibble()
IL6R_T1D    <- import("data/cis_sumstats/IL6R_T1D.rds")  %>% as_tibble()
IL2RB_T1D   <- import("data/cis_sumstats/IL2RB_T1D.rds") %>% as_tibble()
IL2RG_T1D   <- import("data/cis_sumstats/IL2RG_T1D.rds") %>% as_tibble()
IL2RA_T1D   <- import("data/cis_sumstats/IL2RA_T1D.rds") %>% as_tibble()
IFNAR2_T1D  <- import("data/cis_sumstats/IFNAR2_T1D.rds") %>% as_tibble()
```

There were no missing rsid values (suppressed)

```{r, include = FALSE}
#every 
IL12B_T1D  %>% pull(hm_rsid) %>% anyNA()
TYK2_T1D   %>% pull(hm_rsid) %>% anyNA() 
JAK1_T1D   %>% pull(hm_rsid) %>% anyNA()
JAK2_T1D   %>% pull(hm_rsid) %>% anyNA()
JAK3_T1D   %>% pull(hm_rsid) %>% anyNA()
IL6ST_T1D  %>% pull(hm_rsid) %>% anyNA()
IL6R_T1D   %>% pull(hm_rsid) %>% anyNA()
IL2RB_T1D  %>% pull(hm_rsid) %>% anyNA()
IL2RG_T1D  %>% pull(hm_rsid) %>% anyNA()
IL2RA_T1D  %>% pull(hm_rsid) %>% anyNA()
IFNAR2_T1D %>% pull(hm_rsid) %>% anyNA()
```

## Remove duplicated SNPs

All data frames had duplicated rsids (suppressed). They are
removed.
```{r, results = 'hide'}
IL12B_T1D  %>% pull(hm_rsid) %>% anyDuplicated()
TYK2_T1D   %>% pull(hm_rsid) %>% anyDuplicated() 
JAK1_T1D   %>% pull(hm_rsid) %>% anyDuplicated()
JAK2_T1D   %>% pull(hm_rsid) %>% anyDuplicated()
JAK3_T1D   %>% pull(hm_rsid) %>% anyDuplicated()
IL6ST_T1D  %>% pull(hm_rsid) %>% anyDuplicated()
IL6R_T1D   %>% pull(hm_rsid) %>% anyDuplicated()
IL2RB_T1D  %>% pull(hm_rsid) %>% anyDuplicated()
IL2RG_T1D  %>% pull(hm_rsid) %>% anyDuplicated()
IL2RA_T1D  %>% pull(hm_rsid) %>% anyDuplicated()
IFNAR2_T1D %>% pull(hm_rsid) %>% anyDuplicated()
```


```{r, results = 'hide'}
duplicated_snps_IL12B  <- IL12B_T1D  %>% 
  filter(duplicated(hm_rsid)) %>% pull(hm_rsid)
duplicated_snps_TYK2   <- TYK2_T1D   %>% 
  filter(duplicated(hm_rsid)) %>% pull(hm_rsid) 
duplicated_snps_JAK1   <- JAK1_T1D   %>% 
  filter(duplicated(hm_rsid)) %>% pull(hm_rsid)
duplicated_snps_JAK2   <- JAK2_T1D   %>% 
  filter(duplicated(hm_rsid)) %>% pull(hm_rsid)
duplicated_snps_JAK3   <- JAK3_T1D   %>% 
  filter(duplicated(hm_rsid)) %>% pull(hm_rsid)
duplicated_snps_IL6ST  <- IL6ST_T1D  %>%
  filter(duplicated(hm_rsid)) %>% pull(hm_rsid)
duplicated_snps_IL6R   <- IL6R_T1D   %>% 
  filter(duplicated(hm_rsid)) %>% pull(hm_rsid)
duplicated_snps_IL2RB  <- IL2RB_T1D  %>% 
  filter(duplicated(hm_rsid)) %>% pull(hm_rsid)
duplicated_snps_IL2RG  <- IL2RG_T1D  %>% 
  filter(duplicated(hm_rsid)) %>% pull(hm_rsid)
duplicated_snps_IL2RA  <- IL2RA_T1D  %>% 
  filter(duplicated(hm_rsid)) %>% pull(hm_rsid)
duplicated_snps_IFNAR2 <- IFNAR2_T1D  %>% 
  filter(duplicated(hm_rsid)) %>% pull(hm_rsid)

IL12B_T1D  <- IL12B_T1D   %>% filter(!hm_rsid %in% duplicated_snps_IL12B)
TYK2_T1D   <- TYK2_T1D    %>% filter(!hm_rsid %in% duplicated_snps_TYK2) 
JAK1_T1D   <- JAK1_T1D    %>% filter(!hm_rsid %in% duplicated_snps_JAK1)
JAK2_T1D   <- JAK2_T1D    %>% filter(!hm_rsid %in% duplicated_snps_JAK2)
JAK3_T1D   <- JAK3_T1D    %>% filter(!hm_rsid %in% duplicated_snps_JAK3)
IL6ST_T1D  <- IL6ST_T1D %>% filter(!hm_rsid %in% duplicated_snps_IL6ST)
IL6R_T1D   <- IL6R_T1D    %>% filter(!hm_rsid %in% duplicated_snps_IL6R)
IL2RB_T1D  <- IL2RB_T1D   %>% filter(!hm_rsid %in% duplicated_snps_IL2RB)
IL2RG_T1D  <- IL2RG_T1D   %>% filter(!hm_rsid %in% duplicated_snps_IL2RG)
IL2RA_T1D  <- IL2RA_T1D   %>% filter(!hm_rsid %in% duplicated_snps_IL2RA)
IFNAR2_T1D <- IFNAR2_T1D  %>% filter(!hm_rsid %in% duplicated_snps_IFNAR2)
```


## Write data in TwoSampleMR format

```{r}

#Change to numeric
rename_for_TwoSampleMR_o  <- function(data, beta, other_allele, effect_allele,
                                      hm_rsid, hm_beta, standard_error, 
                                      hm_effect_allele_frequency,
                                      hm_effectAllele, hm_otherAllele,
                                      p_value, hm_pos) {
  data %>%
  select(-beta, -other_allele, -effect_allele) %>%
  rename(SNP = hm_rsid, beta = hm_beta, se = standard_error, eaf = hm_effect_allele_frequency,
         effect_allele = hm_effect_allele, other_allele = hm_other_allele,
         pval = p_value, pos = hm_pos) %>%
  mutate(Phenotype = "Risk of type 1 diabetes",
         ncase = 18942, ncontrol = 501638, samplesize = 18942+501638)
  }

IL12B_T1D   <- IL12B_T1D  %>% rename_for_TwoSampleMR_o()
IL2RA_T1D   <- IL2RA_T1D  %>% rename_for_TwoSampleMR_o()
IL2RB_T1D   <- IL2RB_T1D  %>% rename_for_TwoSampleMR_o()
IL2RG_T1D   <- IL2RG_T1D  %>% rename_for_TwoSampleMR_o()
IL6ST_T1D   <- IL6ST_T1D  %>% rename_for_TwoSampleMR_o()
IL6R_T1D    <- IL6R_T1D   %>% rename_for_TwoSampleMR_o()
JAK1_T1D    <- JAK1_T1D   %>% rename_for_TwoSampleMR_o()
JAK2_T1D    <- JAK2_T1D   %>% rename_for_TwoSampleMR_o()
JAK3_T1D    <- JAK3_T1D   %>% rename_for_TwoSampleMR_o()
TYK2_T1D    <- TYK2_T1D   %>% rename_for_TwoSampleMR_o()
IFNAR2_T1D  <- IFNAR2_T1D %>% rename_for_TwoSampleMR_o()
```


Write data in TwoSampleMR data format
```{r}
IL12B_T1D  %>% 
  export("data/export/IL12B_T1D_TwoSampleMR.csv")
IL2RA_T1D  %>% 
  export("data/export/IL2RA_T1D_TwoSampleMR.csv")
IL2RB_T1D  %>% 
  export("data/export/IL2RB_T1D_TwoSampleMR.csv")
IL2RG_T1D  %>% 
  export("data/export/IL2RG_T1D_TwoSampleMR.csv")
IL6ST_T1D  %>%
  export("data/export/IL6ST_T1D_TwoSampleMR.csv")
IL6R_T1D  %>%
  export("data/export/IL6R_T1D_TwoSampleMR.csv")
JAK1_T1D   %>% 
  export("data/export/JAK1_T1D_TwoSampleMR.csv")
JAK2_T1D   %>% 
  export("data/export/JAK2_T1D_TwoSampleMR.csv")
JAK3_T1D   %>% 
  export("data/export/JAK3_T1D_TwoSampleMR.csv")
TYK2_T1D   %>% 
  export("data/export/TYK2_T1D_TwoSampleMR.csv")
IFNAR2_T1D   %>% 
  export("data/export/IFNAR2_T1D_TwoSampleMR.csv")
```

```{r, results='hide'}
#Sanity check: no warnings

# read_outcome_data("data/export/IL12B_T1D_TwoSampleMR.csv", sep = ",")
# read_outcome_data("data/export/IL2RA_T1D_TwoSampleMR.csv", sep = ",")
# read_outcome_data("data/export/IL2RB_T1D_TwoSampleMR.csv", sep = ",")
# read_outcome_data("data/export/IL2RG_T1D_TwoSampleMR.csv", sep = ",")
# read_outcome_data("data/export/IL6ST_T1D_TwoSampleMR.csv", sep = ",")
# read_outcome_data("data/export/IL6R_T1D_TwoSampleMR.csv", sep = ",")
# read_outcome_data("data/export/JAK1_T1D_TwoSampleMR.csv", sep = ",")
# read_outcome_data("data/export/JAK2_T1D_TwoSampleMR.csv", sep = ",")
# read_outcome_data("data/export/JAK3_T1D_TwoSampleMR.csv", sep = ",")
# read_outcome_data("data/export/TYK2_T1D_TwoSampleMR.csv", sep = ",")
# read_outcome_data("data/export/IFNAR2_T1D_TwoSampleMR.csv", sep = ",")
```

